---
title: "Homework7"
author: "Jun Rao"
date: "9/21/2020"
output: word_document
---
Homework 7: Dirichlet Process Mixture Models
The goal of this homework is to use the R package dirichletprocess to implement an infinite dimensional Gaussian mixture model, then compare it to finite Gaussian mixtures with model selection via cross-validation (mclust).

1.First use rnorm to generate random data from 5 cluster centers, as we did in class last week (use 20 data points per cluster center). Scale the data set. 
```{r message=FALSE}
library(ggplot2)
library(data.table)
library(mclust)
library(dirichletprocess)
N.true.clusters <- 5
true.means <- 1:N.true.clusters
N.simulated.data <- 20*N.true.clusters
means.simulated.data <- rep(true.means, each=N.simulated.data/N.true.clusters)

# set.seed(1)
true.sd <- 0.2
sim.data.vec <- rnorm(N.simulated.data, means.simulated.data, sd=true.sd)

## linear transformation on each column so that mean=0 and sd=1.
df <- scale(sim.data.vec)
```
Then use dirichletprocess::DirichletProcessGaussian to fit the mixture model,  Use system.time on the model fitting process – how long does it take?
```{r }

dp <- DirichletProcessGaussian(df)

system.time(dp<- Fit(dp, its=100))#iterations of sampling algorithm.

```
It takes 5.32 seconds
then use the plot function to make a figure similar to Figure 1 (right) in vignette(“dirichletprocess”, package=”dirichletprocess”) (histogram for data, solid/dashed lines for the model).
```{r }
par(mfrow=c(1,2))
plot(dp)
plot(dp, data_method="hist")
```
2.Now use mclust to fit a model with 5 clusters to the scaled data. Make a plot similar to the previous one. Use geom_bar or geom_histogram to plot the data. Use the mclust::dens function to compute the mixture density to plot with geom_line. Do the two models look similar?
```{r }
#Now use mclust to fit a model with 5 clusters to the scaled data.
fit <- Mclust(df,5)
conc.vec <- seq(min(df),max(df),l=100)

#Use the mclust::dens function to compute the mixture density
Dens <- dens(modelName = fit$modelName,data = conc.vec,parameters = fit$parameters)

density <- data.table(concentration = conc.vec,den = Dens)

ggplot() + geom_histogram(aes(x=df, y =stat(count) /10)) + geom_line(aes(concentration,den),data=density,color='red')
```

These two models doesn't simlier to each other
3.Divide the data into 50% train, 50% validation. Use mclust to fit a mixture model with K=1 to 10 components on the train set, then compute and plot y=validation negative log likelihood as a function of x=number of components. Use system.time on the whole process (excluding the plot) – how much time does it take to do model selection yourself? Is that faster or slower than the Dirichlet Process mixture model from question?

```{r }
set.prop.vec <- c(validation=0.5, train=0.5)
N <- length(df)
rounded.counts <- floor(set.prop.vec*(N))
not.shuffled.sets <- rep(names(set.prop.vec), rounded.counts)

set.seed(1)
shuffled.sets <- sample(not.shuffled.sets)
clusters <- 10

lik.dt.list <- list()


system.time(
  for(n.clusters in 1:clusters){
    for (set in names(set.prop.vec)) {
        data.set <- df[shuffled.sets == set,]
        mclust <- Mclust(data.set, n.clusters, modelNames="E")
        log.lik.vec <- dens(
                             modelName = mclust[["modelName"]],
                             data = data.set,
                             parameters = mclust[["parameters"]],
                             logarithm=TRUE)
       total.log.lik <- sum(log.lik.vec)
       rbind(my=total.log.lik, mclust=mclust[["loglik"]])
       lik.dt.list[[paste(n.clusters, set)]] <- data.table(n.clusters, set, total.log.lik)
    }

}
)

lik.dt <- do.call(rbind, lik.dt.list)

ggplot()+
  geom_line(aes(
    n.clusters, -total.log.lik, color=set),
    data=lik.dt)
```


It takes 1.20 seconds. It is much faster than the Dirichlet Process mixture model


