---
title: "Homework11"
author: "Jun Rao"
date: "10/26/2020"
output: word_document
---
Segmentation model selection and evaluation

The goals of this homework are to
explore how a subtrain/validation split can be used to select the segmentation model size parameter.
compare segmentation models in terms of label error rates and ROC curves
```{r}
library(ggplot2)
library(data.table)
data(neuroblastoma, package="neuroblastoma")
```

1.For data sequence profile.id=79, chromosome=2 in the neuroblastoma data, assign observations to 50% subtrain 50% validation sets. How many observations are in each set? Hint: use table(set)


```{r}
pro.dt <- data.table(neuroblastoma$profiles)[profile.id=="79" & chromosome=="2"]
pro.dt[, set := rep(rep(c("subtrain","validation"), each=2), l=.N)]
pro.dt[, index := 1:.N]

subtrain.dt <- pro.dt[set=="subtrain"]
validation.dt <- pro.dt[set== "validation"]

str(subtrain.dt)

```

2.Use binsegRcpp::binseg_normal to compute models from 1 to 10 segments on the subtrain data, then compute one data table with predicted changepoint positions

```{r}

subtrain.binseg.models <- binsegRcpp::binseg_normal(subtrain.dt$logratio, 10)
subtrain.binseg.dt <- coef(subtrain.binseg.models)

subtrain.getPos <- function(subtrain.index){
    orig.index <- subtrain.dt$index[ subtrain.index ]
    pro.dt$position[ orig.index ]
}


subtrain.binseg.dt[, start_in_all_data := subtrain.getPos(subtrain.binseg.dt$start)]
subtrain.binseg.dt[, end_in_all_data := subtrain.getPos(subtrain.binseg.dt$end)]
subtrain.binseg.dt[, type := rep("subtrain", l=.N)]
head(subtrain.binseg.dt)
```
 and another data table with segment start/end values (in terms of the full data indices/positions, not just the subtrain set). 
```{r}

validation.binseg.models <- binsegRcpp::binseg_normal(validation.dt$logratio, 10)
validation.binseg.dt <- coef(validation.binseg.models)

validation.getPos <- function(validation.index){
    orig.index <- validation.dt$index[ validation.index ]
    pro.dt$position[orig.index ]
}


validation.binseg.dt[, start_in_all_data := validation.getPos(validation.binseg.dt$start)]
validation.binseg.dt[, end_in_all_data := validation.getPos(validation.binseg.dt$end)]
validation.binseg.dt[, type := rep("validation", l=.N)]
head(validation.binseg.dt)

```
Plot these models on top of the data (use points with different color/fill for different sets, black=subtrain, red=validation), and use facet_grid(segments ~ .) to draw each model size in a different panel. After how many segments does the model appear to overfit?

```{r}
total.dt <- rbind(subtrain.binseg.dt,validation.binseg.dt)

for(col.name in c("start_in_all_data", "end_in_all_data")){
  col.value <- total.dt[[col.name]]
  set(total.dt, j=paste0(col.name, ".pos"),
  value=pro.dt$position[col.value])
}

total.dt[, end.before := c(NA, end_in_all_data[-.N]), by=.(type, segments) ]
change.dt <- data.table(pro.dt, total.dt[1 < start_in_all_data])
change.dt[, changepoint := (start_in_all_data+end.before)/2]


gg <- ggplot(aes(x=position, y=logratio), data=pro.dt)+geom_point()

(gg+ facet_grid(segments ~ .)+
  geom_segment(aes(
    x=start, y=mean,
    xend=end, yend=mean,
    color=type),
    data=total.dt) + 
  geom_vline(aes(
    xintercept=end_in_all_data,
    color=type),
    data=change.dt)+
  scale_size_manual(values=c(subtrain=2, validation=1))
)




```
 
3.Compute the square loss for each model size and set, and store these numbers in a data table with columns segments,set,loss. Plot the loss as a function of the number of segments, with different sets in different colors (black=subtrain, red=validation). Based on the loss values/plot, what is the number of segments you should select?
```{r}
subtrain.loss.dt<- data.table()
validation.loss.dt <- data.table()
subtrain.loss.dt[, loss := subtrain.binseg.models$loss]
subtrain.loss.dt[, segments := subtrain.binseg.models$segments]
subtrain.loss.dt[, type := "subtrain"]
validation.loss.dt[, loss := validation.binseg.models$loss]
validation.loss.dt[, segments := validation.binseg.models$segments]
validation.loss.dt[, type := "validation"]
loss.dt <- rbind(subtrain.loss.dt,validation.loss.dt)


ggplot()+
  geom_line(aes(
    segments, loss, color=type),
    data=loss.dt)+
  xlab("the number of segments") + ylab("the loss value")
```
4.Compute model selection criteria = loss + penalty*segments, for penalty=2 (AIC) and penalty=log(N.data) (BIC). Save these numbers in a single data table with columns segments, crit.name(AIC/BIC), crit.value(numeric value defined in equation above). Then plot them using different colors for different penalties (e.g., AIC=blue, BIC=orange). Do the two criteria select the same number of segments?
```{r}
binseg.models <- binsegRcpp::binseg_normal(pro.dt$logratio, 10)
AIC.criteria.dt<- data.table()
BIC.criteria.dt <- data.table()

AIC.criteria.dt[, criteria.value := binseg.models$loss + 2 * binseg.models$segments]
AIC.criteria.dt[, segments := binseg.models$segments]
AIC.criteria.dt[, criteria.name := "AIC"]

BIC.criteria.dt[, criteria.value := binseg.models$loss + log(length(pro.dt$logratio)) * binseg.models$segments]
BIC.criteria.dt[, segments := binseg.models$segments]
BIC.criteria.dt[, criteria.name := "BIC"]

criteria.dt <- rbind(AIC.criteria.dt,BIC.criteria.dt)


ggplot()+
  geom_line(aes(
    segments, criteria.value, color=criteria.name),
    data=criteria.dt)+
  xlab("the number of segments") + ylab("the criteria value")
```

The goal is to run binsegRcpp::binseg_normal on ALL labeled neuroblastoma data sequences, and compare two model selection criteria (AIC/BIC) in terms of label error rates and ROC curves.
You should have a for loop over rows of the “annotations” data table, since there is only one label/row per data sequence.




